---
title: "Homework 5"
author: "Taylor Evans"
date: "2026-09-29"
output: pdf_document
---
## Set up

```{r}
library(palmerpenguins)
head(penguins)
```

-  `library(palmerpenguins)` loads the package, which makes the dataset available as `penguins`.
-  `head(penguins)` shows the first six rows so we can inspect the variables


## Task 1: Calculate Summary Statistics
*Objective: Use the 'apply' function to calculate column-wise means for the numeric variables in the dataset.*

```{r}
numeric_means <- apply(penguins[, sapply(penguins, is.numeric)], 2, mean, na.rm = TRUE)
print(numeric_means)
```

The `apply()` function evaluates a function over the margins of an array or matrix. Here it is used to compute the mean of each numeric column in the dataset.

-  penguins[, sapply(penguins, is.numeric)]` selects only the numeric columns from the dataset. `sapply` checks each column with `is.numeric` and returns TRUE or FALSE, and that result keeps only the numeric columns.
-  The `2` is the MARGIN argument. It specifies that the function should be applied to columns (use `1` for rows). MARGIN 2 preserves the columns and collapses the rows.
-  `mean, na.rm = TRUE` calculates the mean for each column, ignoring missing (`NA`) values. `na.rm = TRUE` is passed to `mean` through the `...` argument.


## Task 2: Count Penguins by Species
*Objective: Use the 'tapply' function to count the number of penguins for each species.*

```{r}
species_counts <- tapply(penguins$species, penguins$species, length)
print(species_counts)
```

The `tapply()` function is short for table apply, and it applies a function over subsets of a vector. Here it is used to count the number of penguins per species.

-  The first `penguins$species` is the vector that is split into groups based on unique species values.
-  The second `penguins$species` specifies the factor to group by.
-  `length` counts the number of occurrences in each group.

## Task 3: Analyze Bill Length by Species
*Objective: Use the 'lapply' function to calculate the mean bill length for each species.*

```{r}
bill_length_means <- lapply(split(penguins$bill_length_mm, penguins$species),
                            mean, na.rm = TRUE)
print(bill_length_means)
```

The `lapply()` function loops over a list of objects and applies a function to every element of that list. Here it is used to calculate the mean bill length for each species. 

-  `split(penguins$bill_length_mm, penguins$species)` splits the bill lengths into separate lists based on species.
-  `lapply(..., mean, na.rm = TRUE)` applies the `mean()` function to each species' list.
-  `na.rm = TRUE` ensures missing values (`NA`) are ignored.


## Task 4: Create a Summary Table
*Objective: Use the sapply function to create a summary table with the mean and standard deviation for each numeric variable.*

```{r}
summary_table <- sapply(penguins[, sapply(penguins, is.numeric)],
                        function(x) c(mean = mean(x, na.rm = TRUE),
                                      sd = sd(x, na.rm = TRUE)))
print(summary_table)
```

The `sapply()` function is a variant of `lapply` that simplifies the result if possible. Here it is used to compute both the mean and standard deviation for each numeric variable.

-  `penguins[, sapply(penguins, is.numeric)]` selects only the numeric columns.
-  The anonymous function `function(x) c(mean = mean(x, na.rm = TRUE), sd = sd(x, na.rm = TRUE))` calculates both the mean and the standard deviation. An anonymous function is a function that does not have a name, it is written on the fly inside the loop function.

